---
title: "Substance-aware unit conversions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Substance-aware unit conversions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(substances)
```

## The gap this fills

`units` converts between commensurable units. It will not convert mg/dL to
mmol/L, and it is right not to: those measure different dimensions, and the
bridge between them is the analyte's molar mass. A unit string has nowhere to
put one.

```{r}
tryCatch(units::set_units(units::set_units(100, "mg/dL"), "mmol/L"),
         error = function(e) conditionMessage(e))
```

Attach the substance and it resolves:

```{r}
set_units(set_substances(100, "glucose", "mg/dL"), "mmol/L")
```

## Why the substance cannot live in the unit

The tempting shortcut is to define per-substance units in udunits. It does not
work, and the way it fails is worth seeing, because it is what shaped the rest
of the design.

Define them relative to `mol` and every substance becomes mutually convertible:

```{r}
units::install_unit("mol_glucose", "0.0055507 mol")
units::install_unit("mol_sodium",  "0.0434978 mol")

units::set_units(1, "mol_glucose") + units::set_units(1, "mol_sodium")
```

A mole of glucose plus a mole of sodium is not 8.84 of anything. Nothing warns,
because as far as udunits is concerned these are two scalings of one dimension.

```{r}
units::remove_unit("mol_glucose"); units::remove_unit("mol_sodium")
```

Define them as *base* units instead and the isolation is perfect, but nothing
converts:

```{r}
units::install_unit("mol_glucose"); units::install_unit("g_glucose")
c(across_substances = units::ud_are_convertible("mol_glucose", "mol"),
  within_substance  = units::ud_are_convertible("mol_glucose", "g_glucose"))
units::remove_unit("mol_glucose"); units::remove_unit("g_glucose")
```

Adding the definition that would make the second `TRUE` is exactly what
re-creates the first. **Inside udunits, isolation and conversion are mutually
exclusive.** So here the substance is a property of the vector, and the registry
supplies the quantities that bridge dimensions `units` cannot relate on its own.

## A different substance per element

Long-format laboratory data has one analyte per row, so that is the ordinary
case rather than an edge case. A `substances` vector is a `units` vector with
one extra attribute -- the substance of every element -- so everything `units`
can do it can do, and the substance survives subsetting, reordering and
data-frame operations.

```{r}
class(set_substances(1, "glucose", "mg/dL"))
```

```{r}
x <- set_substances(c(100, 140, 90), c("glucose", "sodium", "glucose"), "mg/dL")
x
substances(x[2:3])

d <- data.frame(id = 1:3, result = x)
d[d$id > 1, ]
```

Every element converts by its own substance, in one call:

```{r}
set_units(x, "mmol/L")
```

Quantities of different substances do not combine:

```{r}
tryCatch(set_substances(1, "glucose", "mmol/L") + set_substances(1, "sodium", "mmol/L"),
         error = function(e) conditionMessage(e))
```

When the unit varies by row too — which is how the data usually arrives — use
`mixed_substances()`. Its job is to become a homogeneous `substances` vector,
which is where arithmetic happens.

```{r}
m <- mixed_substances(c(100, 5.5, 140),
                      c("mg/dL", "mmol/L", "mg/dL"),
                      c("glucose", "glucose", "sodium"))
m
set_units(m, "mmol/L")
```

## Conversions are derived, not tabulated

A substance carries parameters that are themselves `units` quantities. A
conversion is the product of small integer powers of them that makes the two
units commensurable. One molar mass therefore serves every unit pair it can
bridge, rather than needing a table row each:

```{r}
set_units(set_substances(1, "glucose", "g"),     "mol")
set_units(set_substances(1, "glucose", "ug/L"),  "nmol/L")
set_units(set_substances(1, "glucose", "mg/dL"), "mmol/L")
```

Where `units` can already do the job, nothing here is involved at all:

```{r}
set_units(set_substances(1, "albumin", "mg/dL"), "g/L")
```

### Parameters compose

Sodium in mEq/L needs the molar mass *and* the valence, and the search finds
both:

```{r}
set_units(set_substances(1, "sodium", "mg/dL"), "meq/L")
```

### Volume: density for condensed phases, molar volume for gases

Which parameter bridges volume depends on the state of the substance, and the
distinction is not cosmetic.

For a **solid or liquid**, density links mass and volume directly:

```{r}
set_units(set_substances(19.3, "gold", "g"), "cm^3")
```

and combining it with the atomic weight gives the molar volume, a two-parameter
path:

```{r}
set_units(set_substances(1, "gold", "mol"), "cm^3")
set_units(set_substances(1, "mercury", "mol"), "cm^3")
```

For a **gas**, that route is the wrong one. Avogadro's law says a mole of any
gas occupies the same volume at the same temperature and pressure, so the bridge
from amount to volume is the molar volume, and it is very nearly not
substance-specific at all:

```{r}
set_units(set_substances(1, "helium", "mol"),     "L")
set_units(set_substances(1, "neon", "mol"),       "L")
set_units(set_substances(1, "dihydrogen", "mol"), "L")
set_units(set_substances(1, "dinitrogen", "mol"), "L")
```

Deriving volume from mass instead would make that look substance-specific, and
it would quietly answer a different question. A tabulated element density is a
property of the element **in its standard state** — which for hydrogen is H~2~ —
while the atomic weight describes a single atom. Multiply one by the other and
you get the volume per mole of *atoms*: 11.2 L for hydrogen, half a molar
volume, and a quantity nobody wants, because monatomic hydrogen is not something
you can have.

So the atomic entries of the diatomic elements carry no volume bridge, and say
why:

```{r, error = TRUE}
set_units(set_substances(1, "hydrogen", "mol"), "L")
```

```{r}
substance_info("hydrogen")
```

Elements that are not naturally or typically handled in elemental form —
technetium, the transuranics — are left out of this entirely rather than
carrying a density nobody can use.

### The element and the substance it comes as

That leaves H and H~2~ as two entries, which is the honest position: they are
different substances with different molar masses, and only one of them is
something you can put in a flask. H~2~ carries both bridges — molar mass as a
molecule, molar volume as a gas:

```{r}
h2 <- set_substances(1, "dihydrogen", "mol")

set_units(h2, "g")   # molar mass, as a molecule
set_units(h2, "L")   # molar volume, as a gas
```

while H carries only the atomic weight, which is what an elemental analysis
reports:

```{r}
set_units(set_substances(1, "hydrogen", "mol"), "g")
```

Converting *between* the two — one mole of H~2~ contains two moles of H — is
deliberately not offered. Once a package converts H~2~ to H it is expected to
convert CO~2~ to C, glucose to carbon, and reactants to products, and that is a
different package. Quantities of different substances simply do not combine:

```{r, error = TRUE}
set_substances(2, "carbon_dioxide", "mol") + set_substances(8, "water", "mol")
```

If a project needs a particular relationship, it registers the substance it
actually measures and converts within it, which is the next section.

### Not everything is a molar mass

Insulin's mU/L to pmol/L is fixed by the WHO biological standard, not by
insulin's mass. The registry records it as an `activity`, and insulin's molar
mass — which it also has — plays no part:

```{r}
set_units(set_substances(1, "insulin", "mIU/L"), "pmol/L")
```

### Not everything is multiplicative

HbA1c is affine. It is stored in the direction NGSP publishes it, with the
inverse derived rather than transcribed:

```{r}
set_units(set_substances(c(6.5, 7.0), "hba1c", "%"), "mmol/mol")
```

`%` and `mmol/mol` are both dimensionless, so `units` would happily convert
between them by a factor of 10 and produce a plausible, wrong number. Registered
conversions are therefore consulted *before* the dimensional path.

## Substance identity is the point

Three sugars share a formula and a molar mass, and are not interchangeable:

```{r}
vapply(c("glucose", "galactose", "fructose"),
       function(s) as.numeric(substance_parameters(s)$molar_mass), numeric(1))

tryCatch(set_substances(1, "glucose", "mmol/L") + set_substances(1, "galactose", "mmol/L"),
         error = function(e) conditionMessage(e))
```

A subtler case: blood urea nitrogen is reported as the mass of *nitrogen*, not
of urea. Treating "BUN" as another name for urea would apply urea's molar mass
and be wrong by a factor of 2.14. They are different substances.

```{r}
set_units(set_substances(1, "BUN", "mg/dL"),  "mmol/L")
set_units(set_substances(1, "urea", "mg/dL"), "mmol/L")
```

Names and synonyms resolve case-insensitively, so study-specific labels reach
the right entry:

```{r}
substance_resolve(c("LDL Cholesterol", "HDL-C", "  GLUCOSE ", "Na", "T4"))
```

## Every value carries a citation

The reference data ships as CSV with a `source_id` on each value, so a reviewer
can check any number without reading code:

```{r}
substance_info("LDL Cholesterol")
```

A `status` column lets the registry record that a conversion is contested,
rather than leaving it out — which reads as "not looked up yet" — or including
it, which reads as "fine".

```{r}
tryCatch(set_units(set_substances(50, "Lp(a)", "mg/dL"), "nmol/L"),
         error = function(e) conditionMessage(e))
```

Lp(a) has no molar mass here because apo(a) isoform size varies between
individuals, so no fixed mass-to-molar factor is valid. Iron's valence is
withheld for a similar reason — it circulates as both Fe^2+^ and Fe^3+^, so mEq
is ambiguous — while its other parameters remain usable:

```{r}
names(substance_parameters("iron"))
```

Molar masses of ordinary molecules are computed from the formula rather than
transcribed, so every one of them is re-derivable:

```{r}
molar_mass_from_formula(c("C6H12O6", "C27H46O", "C15H11I4NO4"))
```

## Arithmetic

Concentration times volume is an amount, still of the same substance. Dividing
two quantities of the same substance cancels it, leaving an ordinary `units`
quantity:

```{r}
conc <- set_substances(2, "glucose", "mmol/L")
conc * units::set_units(3, "L")

set_substances(6, "glucose", "mmol/L") / set_substances(2, "glucose", "mmol/L")
```

This interoperation needs R >= 4.3.0, for `chooseOpsMethod()`. Below that R
cannot choose between the `substances` and `units` operator methods: it warns
"Incompatible methods" and falls back to the internal default, which returns the
values with the left operand's unit unchanged — so `2 mmol/L * 3 L` comes back
as `6 mmol/L`. A wrong answer rather than an error is not something a package
about not confusing quantities can ship, which is what settles the version
requirement.

Only `*` and `/` are defined against a bare `units` quantity: adding something
with no substance to something with one is refused rather than guessed at.

The rest of the vector interface is ordinary base R. The reductions — `sum()`,
`mean()`, `median()`, `range()`, `diff()` — insist on a single substance,
because collapsing several elements into one would otherwise average across
analytes:

```{r, error = TRUE}
sum(set_substances(c(100, 140), "glucose", "mg/dL"))
sum(set_substances(c(100, 140), c("glucose", "sodium"), "mg/dL"))
```

`unique()` needs no such rule, because it compares the value *and* the
substance: 5 mmol/L of glucose is not 5 mmol/L of urea.

```{r}
unique(set_substances(c(5, 5, 5), c("glucose", "glucose", "urea"), "mmol/L"))
```

`vctrs` and `pillar` are supported but not required. Where they are installed,
their methods are registered at load time, so `vec_slice()`, `vec_c()` and
`dplyr::bind_rows()` carry the substance too; where they are not, everything
above still works.

## Bringing your own substances

The registry here covers common chemistry and a working clinical set. It is not
meant to be exhaustive, and it is not the only registry you can have. A package
or a project supplies its own with a data frame:

```{r}
my_substances <- data.frame(
  substance_id = c("widgetol", "widgetol", "sprocketase"),
  parameter    = c("molar_mass", "density", "molar_mass"),
  value        = c(100, 1.2, 250),
  unit         = c("g/mol", "g/mL", "g/mol"),
  source_id    = "internal-spec")

substance_system("my_project", inherit = "substances",
                 parameters = my_substances)

set_units(set_substances(1, "widgetol", "mg/dL", system = "my_project"), "mmol/L")
set_units(set_substances(120, "widgetol", "g", system = "my_project"), "mL")
```

There is no identity table in that call. One is derived from the
`substance_id`s the parameters mention, so registering a substance is a row, not
a schema. Supply `substances` explicitly when you want names, formulae and
identifiers as well — the shipped registry does — and the referential check will
then hold you to it.

`inherit = "substances"` keeps everything above and adds to it; omit it for a
registry containing only your own entries. Either way the default registry is
untouched, and vectors from different systems refuse to combine, so one
project's definitions cannot leak into another's session:

```{r, error = TRUE}
set_substances(1, "widgetol", "mg/dL")
```

Synonyms and conversions extend the same way: a `synonyms` data frame maps your
study's labels onto substance ids, and a `conversions` data frame adds affine or
factor relationships that no parameter can derive. Nothing about that machinery
is specific to the substances shipped here.

### Parameter kinds of your own

The bridge parameters are not a closed list either. Nothing in the conversion
machinery knows what a molar mass *is* — it multiplies by whatever quantities a
substance carries and asks `units` which combination lands on the target. A new
kind therefore needs no code, only a declaration of what units it must have.

Enzyme specific activity, for instance, bridges the mass of an enzyme and its
catalytic activity:

```{r}
substance_system("enzymes", parameter_units = c(specific_activity = "U/mg"),
  parameters = data.frame(
    substance_id = "alkaline_phosphatase",
    parameter    = "specific_activity",
    value        = 1000,
    unit         = "U/mg",
    source_id    = "supplier-certificate"))

set_units(set_substances(1, "alkaline_phosphatase", "ug", system = "enzymes"), "U")
```

or a turnover number, which relates enzyme *amount* to activity:

```{r}
substance_system("kinetics", parameter_units = c(turnover = "kat/mol"),
  parameters = data.frame(
    substance_id = "carbonic_anhydrase",
    parameter    = "turnover",
    value        = 1e6,
    unit         = "1/s",
    source_id    = "textbook"))

set_units(set_substances(1, "carbonic_anhydrase", "nmol", system = "kinetics"),
          "mmol/s")
```

`U` and `IU` are both installed at load, and they are deliberately different
dimensions: `U` is the enzyme unit, `umol/min`, while `IU` is the WHO
biological-standard unit whose relationship to mass or amount is fixed by an
assay standard and differs per substance. That is why insulin needs an
`activity` parameter and alkaline phosphatase does not.

The units are checked against the kind when the system is built, so a parameter
that could not possibly bridge what it claims to is caught at registration
rather than at conversion time:

```{r, error = TRUE}
substance_system("wrong", parameters = data.frame(
  substance_id = "x", parameter = "molar_mass", value = 1, unit = "g/L"))
```

## What is out of scope

Normalising unit **strings** — `ng/ml` versus `ng/mL`, `IU/L` versus `U/L`,
`frac of 1` versus `fraction of 1` — is per-source data cleaning and belongs in
the consuming package. Unrecognised unit strings are rejected rather than
guessed at:

```{r}
tryCatch(mixed_substances(1, "frac of 1", "glucose"),
         error = function(e) conditionMessage(e))
```

Conversions `units` already performs are also out of scope. `mg/dL` to `g/L` is
dimensional. So is enzyme activity, once `U` is defined as µmol/min — which this
package does at load, because udunits lacks it:

```{r}
units::set_units(units::set_units(1, "U/L"), "ukat/L")
```

That conversion is the same for every enzyme, so it needs no substance and this
package adds nothing to it. Note the contrast with `IU` in the WHO sense, as
used for insulin above: same string in clinical data, genuinely different
things, which is why they are separate dimensions here.
