---
title: "Substance-aware unit conversions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Substance-aware unit conversions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(substances)
```

## The gap this fills

`units` converts between commensurable units. It will not convert mg/dL to
mmol/L, and it is right not to: those measure different dimensions, and the
bridge between them is the analyte's molar mass. A unit string has nowhere to
put one.

```{r}
tryCatch(units::set_units(units::set_units(100, "mg/dL"), "mmol/L"),
         error = function(e) conditionMessage(e))
```

Attach the substance and it resolves:

```{r}
set_units(substance(100, "mg/dL", "glucose"), "mmol/L")
```

## Why the substance cannot live in the unit

The tempting shortcut is to define per-substance units in udunits. It does not
work, and the way it fails is worth seeing, because it is what shaped the rest
of the design.

Define them relative to `mol` and every substance becomes mutually convertible:

```{r}
units::install_unit("mol_glucose", "0.0055507 mol")
units::install_unit("mol_sodium",  "0.0434978 mol")

units::set_units(1, "mol_glucose") + units::set_units(1, "mol_sodium")
```

A mole of glucose plus a mole of sodium is not 8.84 of anything. Nothing warns,
because as far as udunits is concerned these are two scalings of one dimension.

```{r}
units::remove_unit("mol_glucose"); units::remove_unit("mol_sodium")
```

Define them as *base* units instead and the isolation is perfect, but nothing
converts:

```{r}
units::install_unit("mol_glucose"); units::install_unit("g_glucose")
c(across_substances = units::ud_are_convertible("mol_glucose", "mol"),
  within_substance  = units::ud_are_convertible("mol_glucose", "g_glucose"))
units::remove_unit("mol_glucose"); units::remove_unit("g_glucose")
```

Adding the definition that would make the second `TRUE` is exactly what
re-creates the first. **Inside udunits, isolation and conversion are mutually
exclusive.** So here the substance is a property of the vector, and the registry
supplies the quantities that bridge dimensions `units` cannot relate on its own.

## A different substance per element

Long-format laboratory data has one analyte per row, so that is the ordinary
case rather than an edge case. The substance is a field of the vector, and it
survives subsetting, reordering and data-frame operations.

```{r}
x <- substance(c(100, 140, 90), "mg/dL", c("glucose", "sodium", "glucose"))
x
substance_of(x[2:3])

d <- data.frame(id = 1:3, result = x)
d[d$id > 1, ]
```

Every element converts by its own substance, in one call:

```{r}
set_units(x, "mmol/L")
```

Quantities of different substances do not combine:

```{r}
tryCatch(substance(1, "mmol/L", "glucose") + substance(1, "mmol/L", "sodium"),
         error = function(e) conditionMessage(e))
```

When the unit varies by row too — which is how the data usually arrives — use
`mixed_substances()`. Its job is to become a homogeneous `substance`, which is
where arithmetic happens.

```{r}
m <- mixed_substances(c(100, 5.5, 140),
                      c("mg/dL", "mmol/L", "mg/dL"),
                      c("glucose", "glucose", "sodium"))
m
set_units(m, "mmol/L")
```

## Conversions are derived, not tabulated

A substance carries parameters that are themselves `units` quantities. A
conversion is the product of small integer powers of them that makes the two
units commensurable. One molar mass therefore serves every unit pair it can
bridge, rather than needing a table row each:

```{r}
set_units(substance(1, "g", "glucose"),     "mol")
set_units(substance(1, "ug/L", "glucose"),  "nmol/L")
set_units(substance(1, "mg/dL", "glucose"), "mmol/L")
```

Where `units` can already do the job, nothing here is involved at all:

```{r}
set_units(substance(1, "mg/dL", "albumin"), "g/L")
```

### Parameters compose

Sodium in mEq/L needs the molar mass *and* the valence, and the search finds
both:

```{r}
set_units(substance(1, "mg/dL", "sodium"), "meq/L")
```

### Density, and the molar volume it implies

Densities of the elements at their standard state are registered, so mass and
volume are bridged too:

```{r}
set_units(substance(19.3, "g", "gold"), "cm^3")
```

Combined with the atomic weight, density gives the molar volume — molar mass
times the reciprocal of density, a two-parameter path:

```{r}
set_units(substance(1, "mol", "gold"), "cm^3")
set_units(substance(1, "mol", "sodium"), "cm^3")
```

Gas densities are recorded at STP, and the arithmetic comes out where it should.
A monatomic gas occupies the full molar volume; a diatomic one occupies half of
it *per mole of atoms*, because the registry's molar mass is atomic while the
density is of the bulk material:

```{r}
set_units(substance(1, "mol", "helium"),   "L")
set_units(substance(1, "mol", "hydrogen"), "L")
```

That is not a fudge: a mole of H atoms is half a mole of H~2~, which is half of
22.4 L. Mass is conserved whatever the molecular form, so a bulk density and an
atomic molar mass stay consistent. The reference conditions matter enormously
for a gas, so they are recorded rather than assumed:

```{r}
substance_info("hydrogen")
```

### Not everything is a molar mass

Insulin's mU/L to pmol/L is fixed by the WHO biological standard, not by
insulin's mass. The registry records it as an `activity`, and insulin's molar
mass — which it also has — plays no part:

```{r}
set_units(substance(1, "mIU/L", "insulin"), "pmol/L")
```

### Not everything is multiplicative

HbA1c is affine. It is stored in the direction NGSP publishes it, with the
inverse derived rather than transcribed:

```{r}
set_units(substance(c(6.5, 7.0), "%", "hba1c"), "mmol/mol")
```

`%` and `mmol/mol` are both dimensionless, so `units` would happily convert
between them by a factor of 10 and produce a plausible, wrong number. Registered
conversions are therefore consulted *before* the dimensional path.

## Substance identity is the point

Three sugars share a formula and a molar mass, and are not interchangeable:

```{r}
vapply(c("glucose", "galactose", "fructose"),
       function(s) as.numeric(substance_parameters(s)$molar_mass), numeric(1))

tryCatch(substance(1, "mmol/L", "glucose") + substance(1, "mmol/L", "galactose"),
         error = function(e) conditionMessage(e))
```

A subtler case: blood urea nitrogen is reported as the mass of *nitrogen*, not
of urea. Treating "BUN" as another name for urea would apply urea's molar mass
and be wrong by a factor of 2.14. They are different substances.

```{r}
set_units(substance(1, "mg/dL", "BUN"),  "mmol/L")
set_units(substance(1, "mg/dL", "urea"), "mmol/L")
```

Names and synonyms resolve case-insensitively, so study-specific labels reach
the right entry:

```{r}
substance_resolve(c("LDL Cholesterol", "HDL-C", "  GLUCOSE ", "Na", "T4"))
```

## Every value carries a citation

The reference data ships as CSV with a `source_id` on each value, so a reviewer
can check any number without reading code:

```{r}
substance_info("LDL Cholesterol")
```

A `status` column lets the registry record that a conversion is contested,
rather than leaving it out — which reads as "not looked up yet" — or including
it, which reads as "fine".

```{r}
tryCatch(set_units(substance(50, "mg/dL", "Lp(a)"), "nmol/L"),
         error = function(e) conditionMessage(e))
```

Lp(a) has no molar mass here because apo(a) isoform size varies between
individuals, so no fixed mass-to-molar factor is valid. Iron's valence is
withheld for a similar reason — it circulates as both Fe^2+^ and Fe^3+^, so mEq
is ambiguous — while its other parameters remain usable:

```{r}
names(substance_parameters("iron"))
```

Molar masses of ordinary molecules are computed from the formula rather than
transcribed, so every one of them is re-derivable:

```{r}
molar_mass_from_formula(c("C6H12O6", "C27H46O", "C15H11I4NO4"))
```

## Arithmetic

Concentration times volume is an amount, still of the same substance. Dividing
two quantities of the same substance cancels it, leaving an ordinary `units`
quantity:

```{r}
conc <- substance(2, "mmol/L", "glucose")
conc * units::set_units(3, "L")

substance(6, "mmol/L", "glucose") / substance(2, "mmol/L", "glucose")
```

This interoperation needs R >= 4.3.0, for `chooseOpsMethod()`. Below that R
refuses to choose between the `substance` and `units` operator methods and the
expression fails before either runs.

Only `*` and `/` are defined against a bare `units` quantity: adding something
with no substance to something with one is refused rather than guessed at.

## Isolated conversion systems

Because the substance lives in R data rather than global udunits state, several
registries coexist in one session. A project can add or override entries without
touching anyone else's session, and vectors from different systems refuse to
combine.

```{r}
substance_system(
  "my_project", inherit = "substances",
  substances = data.frame(substance_id = "widgetol", name = "Widgetol"),
  parameters = data.frame(substance_id = "widgetol", parameter = "molar_mass",
                          value = 100, unit = "g/mol", status = "ok"))

set_units(substance(1, "mg/dL", "widgetol", system = "my_project"), "mmol/L")

# the default registry is untouched
tryCatch(substance(1, "mg/dL", "widgetol"),
         error = function(e) conditionMessage(e))
```

## What is out of scope

Normalising unit **strings** — `ng/ml` versus `ng/mL`, `IU/L` versus `U/L`,
`frac of 1` versus `fraction of 1` — is per-source data cleaning and belongs in
the consuming package. Unrecognised unit strings are rejected rather than
guessed at:

```{r}
tryCatch(mixed_substances(1, "frac of 1", "glucose"),
         error = function(e) conditionMessage(e))
```

Conversions `units` already performs are also out of scope. `mg/dL` to `g/L` is
dimensional. So is enzyme activity, once `U` is defined as µmol/min — which this
package does at load, because udunits lacks it:

```{r}
units::set_units(units::set_units(1, "U/L"), "ukat/L")
```

That conversion is the same for every enzyme, so it needs no substance and this
package adds nothing to it. Note the contrast with `IU` in the WHO sense, as
used for insulin above: same string in clinical data, genuinely different
things, which is why they are separate dimensions here.
